Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-12-02 Manuscript/TEM_draft.tex
Updated manuscript and ref
reframed discussion + minor changes in intro/methods/results

\newpage
\begin{abstract}
Living species represent a marginal part of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology. This fact has led to an increasing consensus among scientists that both living and fossil taxa must be included in macroevolutionary studies. One approach, the Total Evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both living and fossil taxa at the tips. Although the Total Evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

@@

We find that, when using a clade conservative metric such as Robinson-Foulds distance, Bayesian method recovers the right topology better than Maximum Likelihood method. However, when using triplets method, there is less significant difference between both methods.

-Overall missing data is doesn't affect topology in Bayesian (around minimum 70\% of topological topology in the worth scenario).

-However, one essential way to improve it would be to code missing living taxa.
\end{abstract}

@@

Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}.  Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}. This has led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments (e.g. \citealp{pyrondivergence2011,ronquista2012,schragocombining2013}). %Not a nice sentence

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty (but see \citealp{spencerefficacy2013}). Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence date are used to time calibrate phylogenies \citep{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013}. However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}.  This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data \citep{pyrondivergence2011,ronquista2012,schragocombining2013}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies. However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data that can be difficult (or impossible) to collect.
The morphological data for living taxa is rarely collected when molecular data is available (e.g. \citealt{O'Leary08022013} vs. \citealt{meredithimpacts2011}), and, for fossil taxa, the scarcity of the fossil record only allow to collect the data available (for example, in vertebrates, the hardest parts of the skeleton; \citealt{sansomfossilization2013}). Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. Missing molecular data has been seen by some authors as an issue because it can, in some part of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing to infer topology and branch length)% NC: define phylogenetic signal here. 
    , especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered genes (i.e. approximatively of the genes that are coded for most of the species; \citealt{rouresite-specific2011}) %NC: what do you mean by highly-covered?
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citeauthor{sansomfossilization2013} \citeyear{sansomfossilization2013} but see \citeauthor{pattinsonphylogeny2014} \citeyear{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold were it starts fucking up the phylogeny) but more the fact that Bayesian fails.

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil species. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference from Total Evidence methods.

%ADD APRIL'S PAPER SOMEWHERE

% NC: Somewhere here you need to flag that you're only looking at topologies and briefly explain why.
In this study, we focused only on topology as one of the two aspects of the phylogenetic signal (topology and branch length). Even though both aspects are equally important, branch topology is the first and most straightforward aspect reflecting phylogenetic signal (i.e. topological changes are discrete opposed to branch length changes are continuous). Also, interestingly, the effect of Total Evidence method has not been formally assessed in previous studies using fixed topology \citep{ronquista2012,schragocombining2013}. %TG: or is that mean for them?

Here we use simulations to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. The molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}. The effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}, therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix:
\begin{enumerate}
\item the proportion of living taxa with no morphological data;
\item the proportion of missing data in the fossil taxa; and
\item the proportion of missing morphological characters for both living and fossil taxa in the matrix.
\end{enumerate} 
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect empirical biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases \citep{sansomfossilization2013}, and the overall number of characters depends on the effort of the people identifying them.

We find that when using a Maximum Likelihood approach, as missing data increases, the likelihood of recovering the correct tree topology decreases. However, even with no missing data, Total Evidence matrices dramatically reduce the performance of Bayesian methods for inferring tree topology. We propose that this drastic difference between Bayesian and Maximum Likelihood methods is due to a flattening of the likelihood landscape caused by the unavoidable amount of missing molecular data for fossil taxa in a Total Evidence matrix. We make suggestions for how best to deal with this issue when inferring phylogenies from Total Evidence matrices.
% NC: You don't want to tell the whole story here. Just a results summary and a quick line of interpretation

@@

% NC: Most of the below was fine but I've tried to condense it a bit as it's currently long even for Syst Biol. The key is to explain everything, but with minimal repetition.

@@

\begin{enumerate}
% NC: I'm not going to check but make sure the brief explanations of the parameters here are the same as in the intro and outline above. Also look at my justifications in the intro too. I think this may need a bit more justification as it's the main part of your simulations.
\item{The proportion of living taxa with no morphological data ($M_{L}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that because of the increasing availability of DNA sequences for living taxa, detailed morphological data is scarce.
\item{The proportion of missing data in the fossil taxa ($M_{F}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record. 
\item{the proportion of missing morphological characters for both living and fossil taxa ($M_{C}$): 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa.
\end{enumerate}

@@

\subsubsection{Normalized Tree Similarity}
We used the Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) to be able to compare the two metrics for any $n$ taxa. This index allows to scale the value of any metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are strictly identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topology: a high score of this index (i.e. towards 1) means that the topology is highly conserved between the two trees; on the opposite, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as much as expected when comparing two random trees.

@@

As it would be expected form the literature, the amount of missing data in the morphological matrix does decrease the ability to recover the right topology, regardless of the parameter, the method or the metric used \citep{rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. However each variable does not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam}). Regarding the conservation of entire clades (i.e. the Ronbinson-Foulds distance), the Bayesian consensus trees outperform the other methods and the amount of missing data in the living taxa ($M_{L}$) decreases the most rapidly the clade conservation when using this method. However, when looking at the position of wild card taxa (i.e. the Triplets distance), the Maximum Likelihood outperforms the other methods but with wider distributions overlap (Fig.~\ref{Fig_Results-paircomp_without}-B).

@@

When looking at the global trend of the Bayesian consensus trees for all the parameters combinations, this method outperforms Maximum Likelihood for the Ronbinson-Foulds distance and plateaus to a minimal modal tree similarity of 0.69 regardless the amount of missing data (Fig.~\ref{Fig_Results-global_perparam} - see also \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). However, regarding the Triplets distance, Bayesian consensus trees seems to perform poorly but with great confidence intervals overlaps (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}). Results for the global trend comparison between the uncertainty methods (Bootstrap and Bayesian posterior tree distribution) are available in the supplementary materials (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

When looking at the comparisons between the methods using the Robinson-Foulds distance, the Bayesian consensus trees have the least overlap with respectively the Bayesian posterior trees and the Bootstrap (Fig.~\ref{Fig_Results-paircomp_without}-A). When regarding at the Triplets distance, the two method that have the most distribution are the Bootstraps and the Bayesian posterior trees (Fig.~\ref{Fig_Results-paircomp_without}-B).

@@

% NC: Discussions usually begin with a paragraph that sums up the results. Then you can move on to discussing caveats, problems, extentions etc. You also need to show how your work fits in with the wider literature. So I think you need to re-write this completely with that in mind. Do the same as we did for the introduction - write a line for each paragraph. This reads like you jumped in without thinking clearly about the structure. It probably doesn't need to have all the same subheadings either.

%§1
%Results: global
%   -Missing data increase decreases topological recovery
%The results of our simulations are in adequacy with recent analysis on the effect of missing data in morphological matrices \citep{rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. Our results are showing that ability to recover the correct topology decreases with the amount of missing data, regardless the metric, the method or the parameters considered. In more details, our results shows however a clear superiority of the Bayesian consensus tree over other methods, at least when considering the Robinson-Foulds distance (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). Also, our results suggest that the missing data for missing taxa parameter ($M_{L}$) shows a slightly higher decrease in the ability to recover the correct topology (Fig.~\ref{Fig_Results-paircomp_within}).
%Results 1
%   -However, methods and metrics are unequal
%Results 2
%   -Additionally, parameters have different effect
%Wrap up results
%   -Thus, Baycon is better with loads of living

%§2
%Results from litterature
%   -Missing data per se is not a problem, problem is in the amount of overlap
%Proposed hypothesis
%   -One can reduce overlap by adding living taxa

%§3
%Discussing the metrics in detail
%   -RF is different than triplets and better.
%However, our results can be contrasted whether one looks at the Robinson-Foulds or the Triplets distance. When using the second one, the confidence interval of our different parameters combinations per methods seems higher (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}) and the probability of overlap between the distribution of each parameter is high (Bhattacharyya Coefficient close to 1 - Fig.~\ref{Fig_Results-paircomp_within}-B). The contrast between the results of the two metrics can be explained by the actual nature of the metrics:
%\begin{enumerate}
%\item{The Robinson-Foulds metric is a conservative tree topology metric, it is more sensible to single taxon displacement because it will count clades as similar only if they are composed of the same number of taxa with the same topologies \citep{RF1981}. When getting closer to the root of the tree, displacement of single taxon makes the clades not being exactly identical any more even if the clade still contains all the other taxa in both trees. This metric illustrates therefore the conservation of clades among the compared trees (i.e. when the normalised Robinson-Foulds distance is close to 1, most of the clades in the two trees are identical).}
%\item{On the other hand, the Triplets method is measuring the position of each taxon towards two other reference taxa \citep{critchlowthe1996}. It will penalise only trees where taxa get removed furthest from their original clade. This metric illustrates therefore the amount of wildcard taxa \citep{kearneyfragmentary2002} between the to compared trees (i.e. when the score is closed to 1, view wildcard taxa are present in the tree).}
%\end{enumerate}
%Therefore, in our study, both metrics are showing two different aspects of tree topology. Also, disregarding their fundamental differences in accounting for topological distance, \citet{kuhnerpractical2014} have recently proposed that the Ronbinson-Foulds distance greatly outperforms the Triples distance in signalling differences in topology.

%§4
%Discussing the results 1 in detail
%   -Why is Baycon better than the others?
%Following \citet{kuhnerpractical2014}'s conclusions on the superiority of the Robinson-Foulds distance, the best method to use for recovering the correct topology in a Total Evidence framework is using the Bayesian consensus tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). In fact, regardless of the amount of missing data, this method always displays the highest normalised Robinson-Foulds score (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). When comparing this the probability of overlap in this method between the other onces, it is counter intuitive to note that the Bayesian consensus tree shows the highest dissimilarity with the Bayesian posterior trees from which the consensus tree is build (Fig.~\ref{Fig_Results-paircomp_within}-A). Note that it is also highly dissimilar to the Bootstraps. However that can be linked to the fact that both Bootstraps and Bayesian posterior trees are measured as 125*1000 random pairwise comparisons instead of just 125 tree comparisons for the Bayesian consensus trees. The fact that the 1000 trees to be compared for each of the 125 pairwise comparisons can add noise to the results for these two methods due to the random factor in the pairwise comparisons (see \hyperref[tree_comparisons]{Methods - Tree comparisons}).

>>>>

\newpage
\begin{abstract}
Living species represent a marginal part of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology. This fact has led to an increasing consensus among scientists that both living and fossil taxa must be included in macroevolutionary studies. One approach called Total Evidence, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both living and fossil taxa at the tips. Although the Total Evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

@@

We find that, when using a clade conservative metric such as Robinson-Foulds distance, Bayesian consensus trees recovers the right topology better than Maximum Likelihood or than a subset of Bayesian posterior tree distributions and this regardless of the amount of missing data (minimal tree similarity of 0.69 in the worst scenario). Regarding the data, good topological recovery is not related to the amount of missing data \textit{per se} but to the amount of data overlap. The two factors that influences the most this data overlap are the parameters 1 (proportion of coded living taxa) and 3 (number of morphological characters).
\end{abstract}

@@

Although most species that have ever lived are now extinct \citep{novacek1992ext,raup1993extinction}, the majority of macroevolutionary studies focus solely on living species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as the timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}. This has led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies; a task that remains difficult despite recent methodological developments \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013}. %Not a nice sentence

Up to now, three main approaches have been used to place both living and fossil taxa into phylogenies. These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny, and in which part of the available fossil data is used (i.e. the age of the fossil only or both its age and morphology). Classical cladistic methods use matrices containing morphological data from both living and fossil taxa and treat each taxon as a tip in the phylogeny. Relationships among the taxa are then inferred using optimality criteria such as maximum parsimony \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species and does not allow use of probabilistic methods for dealing with phylogenetic uncertainty. Neontologists, on the other hand, more commonly use probabilistic approaches (e.g. Maximum Likelihood or Bayesian methods) based on matrices containing only molecular data from living species. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these phylogenies and their occurrence date are used to time calibrate phylogenies \citep{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013,wrightbayesian2014}. However neither approach uses all the available data.

A final approach, known as the Total Evidence method, uses matrices containing molecular data from living taxa and morphological data from both living and fossil taxa \citep{eernissetaxonomic1993}. This approach treats every taxon as a tip in the phylogeny, uses the occurrence age of the fossils to time calibrate the phylogeny, and allows the use of probabilistic methods for estimating phylogenetic uncertainty \citep{ronquista2012}. Total Evidence methods have been successfully applied to empirical data \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}, and are becoming an increasingly popular way of adding fossil taxa to phylogenies. However, although the Total Evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data that can be difficult (or impossible) to collect. The morphological data for living taxa is rarely collected when molecular data is available (e.g. \citealt{O'Leary08022013} \textit{vs.} \citealt{meredithimpacts2011}), and, for fossil taxa, the scarcity of the fossil record only allow to collect the data available (for example, in vertebrates, the hardest parts of the skeleton; \citealt{sansomfossilization2013}). Therefore Total Evidence matrices are likely to contain a lot of missing data that may affect the method's ability to infer correct topologies, branch lengths and support values \citep{salamin2003}.

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,sansombias2014}. Missing molecular data has been seen by some authors as an issue because it can, in some part of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing to infer topology and branch length), especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by: (i) including a "modest" number (~50) of highly-covered genes (i.e. a number of genes that are available for all taxa ; \citealt{rouresite-specific2011}); (ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citealt{sansomfossilization2013} but see \citealt{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold were it starts fucking up the phylogeny) but more the fact that Bayesian fails.

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil species. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference from Total Evidence methods. % NC: Somewhere here you need to flag that you're only looking at topologies and briefly explain why.
In this study, we are interested in assessing the effect of missing data on the ability to recover a correct topology. Even though branch length is a second aspect of phylogenetic inferences equally important, topology is the first and most straightforward aspect reflecting phylogenetic signal (i.e. topological changes are discrete opposed to branch length changes are continuous). Also, interestingly, the effect of Total Evidence method has not been formally assessed in previous studies using fixed topology \citep{ronquista2012,schragocombining2013}.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix:
\begin{enumerate}
\item the proportion of living taxa with no morphological data;
\item the proportion of missing data in the fossil taxa; and
\item the proportion of missing morphological characters for both living and fossil taxa in the matrix.
\end{enumerate} 
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect empirical biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases \citep{sansomfossilization2013}, and the overall number of characters depends on the effort of the people identifying them \citep[e.g.][]{O'Leary08022013}.

We find that the ability of recovering the correct topology can mainly be increased when using the Bayesian methods. Also minimizing the proportion of living taxa with no morphological data and the proportion of missing morphological characters improves the ability to recover the correct more than minimizing the proportion of missing data in the fossil record.

@@

@@

\begin{enumerate}
\item{The proportion of living taxa with no morphological data ($M_{L}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that because of the increasing availability of DNA sequences for living taxa, detailed morphological data is scarce.
\item{The proportion of missing data in the fossil taxa ($M_{F}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record. 
\item{the proportion of missing morphological characters for both living and fossil taxa ($M_{C}$): 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa.
\end{enumerate}

@@

\subsubsection{Normalized Tree Similarity}
We used the Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) to be able to compare the two metrics for any $n$ taxa. This index allows to scale the value of any metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are strictly identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topology: a high score of this index (i.e. towards 1) means that the topology is highly conserved between the two trees; on the opposite, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as much as expected when comparing two random trees. Note that both tree comparisons are representing different aspects of topological similarity and do not perform equaly (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

@@

As it would be expected form the literature, the amount of missing data in the morphological matrix does decrease the ability to recover the right topology, regardless of the parameter, the method or the metric used \citep[e.g.][]{rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014}. However each variable does not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam}). Regarding the conservation of entire clades (i.e. the Ronbinson-Foulds distance), the Bayesian consensus trees outperform the other methods and the amount of missing data in the living taxa ($M_{L}$) decreases the most rapidly the clade conservation when using this method. However, when looking at the position of wild card taxa (i.e. the Triplets distance), the Maximum Likelihood outperforms the other methods but with wider distributions overlap (Fig.~\ref{Fig_Results-paircomp_without}-B).

@@

When looking at the global trend of the Bayesian consensus trees for all the parameters combinations, this method outperforms Maximum Likelihood for the Ronbinson-Foulds distance and plateaus to a minimal modal tree similarity of 0.69 regardless the amount of missing data (Fig.~\ref{Fig_Results-global_perparam} - see also \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). However, regarding the Triplets distance, Bayesian consensus trees seems to perform poorly but with great confidence intervals overlaps (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}). These greater larger confidence intervals are due to the lower ability of the Triplet method to highlight precise differences in topology \citep{kuhnerpractical2014}. Results for the global trend comparison between the uncertainty methods (Bootstrap and Bayesian posterior tree distribution) are available in the supplementary materials (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

When looking at the comparisons between the methods using the Robinson-Foulds distance, the Bayesian consensus trees have the least overlap with respectively the Bayesian posterior trees and the Bootstrap (Fig.~\ref{Fig_Results-paircomp_without}-A). When regarding at the Triplets distance, the two method that have the most distribution are the Bootstraps and the Bayesian posterior trees (Fig.~\ref{Fig_Results-paircomp_without}-B). These results are due to the slight difference in comparing topologies between Maximum Likelihood and Bayesian consensus trees and Bootstraps and Bayesian posterior trees: the two last ones are based on a random pairwise comparison of a sub-sample of 1000 trees. This is likely to add noise to the data.

@@

% NC: Discussions usually begin with a paragraph that sums up the results. Then you can move on to discussing caveats, problems, extensions etc. You also need to show how your work fits in with the wider literature. So I think you need to re-write this completely with that in mind. Do the same as we did for the introduction - write a line for each paragraph. This reads like you jumped in without thinking clearly about the structure. It probably doesn't need to have all the same subheadings either.

%§1
%Results: global
%   -Missing data increase decreases topological recovery
The results of our simulations are in adequacy with recent analysis on the effect of missing data in morphological matrices \citep{pattinsonphylogeny2014,wrightbayesian2014,sansombias2014}. Our results are showing that ability to recover the correct topology decreases with the amount of missing data, regardless the metric, the method or the parameters considered.
%Results 1
%   -However, methods and metrics are unequal
However, the effect of missing data on topological recovery varies with different methods to infer topology and different metrics to compare topologies.
%Results 2
%   -Additionally, parameters have different effect
Additionally, within a method and a metric, the parameters used to remove data have different effect on the ability to recover the correct topology.
%Wrap up results
%   -Thus, Baycon is better with loads of living
Thus, we found that using the topologies of Bayesian consensus trees performed consistently better than other methods for recovering the correct topology (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}) and that both the amount of available characters ($M_{C}$) and the amount of data for the living taxa ($M_{L}$) are crucial missing data parameters.

%§2
%Results from litterature
%   -Missing data per se is not a problem, problem is in the amount of overlap
As proposed in numerous previous studies, when recovering topology, the amount of missing data \textit{per se} is not a problem as long as enough data overlaps in the matrix \citep[e.g.][]{rouresite-specific2011,pattinsonphylogeny2014}.
%Proposed hypothesis
%   -One can increase overlap by adding living taxa
To improve topological recovery in a Total Evidence framework, one can efficiently increase the amount of overlapping data by sampling a maximum of characters and by increasing the number of living data coded in the morphological part of the matrix.

%§3
%Discussing the hypothesis 2 in details
%   -Which parameters does what?
For each missing data parameter used in this study, there is a different way to reduce the amount of missing data:
\begin{enumerate}
\item{$M_{L}$:}
One should put more effort in using natural museums history collections for coding the missing morphological data for living taxa if possible
\item{$M_{F}$:}
the amount of missing data unfortunately depends on the quality of the fossil record and can not be actively improved and depends on exceptional discoveries \citep[e.g.][]{nithe2013}
\item{$M_{C}$:}
improvement can be done by vast collaborative projects in order to gather as much characters as possible \citep[e.g.][]{O'Leary08022013}. Therefore, we advocate the importance of coding morphological characters for the most living taxa possible and with the most characters possible, potentially by using collaborative projects portals such as morphobank \citep{morphobank}.
\end{enumerate}
%   -Why is ML more important than others?
Our results are consistent with previous works and are showing that the amount of missing data \textit{per se} does not reduce the ability to recover the correct topology \citep[e.g.][]{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}. In fact, as shown previously, the amount of overlap in the matrix is more crucial \citep[e.g.][]{rouresite-specific2011}. In our case, the overlap in data is greatly reduced when different parameters are combined with a high amount of missing data (e.g. $M_{L}$=75\%, $M_{F}$=75\% and $M_{C}$=75\% - Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). In his scenario the overall number of missing characters $M_{C}$ affects the proportion of overlapping data by reducing the overall size of the matrix (Fig.~\ref{Fig_Results-global_perparam}). However it is important to note that in the pairwise Bhattacharyya Coefficient analysis, the area with the lowest probability of distribution overlap is led by comparing the trees with no missing data for the living taxa to the trees with 75\% missing data for the living data (Right lower corner in Fig.~\ref{Fig_Results-paircomp_within}).

%§4
%Discussing the results 1 in detail
%   -Why is Baycon better than the others?
Following \citet{kuhnerpractical2014}'s conclusions on the superiority of the Robinson-Foulds distance, the best method to use for recovering the correct topology in a Total Evidence framework is using the Bayesian consensus tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). In fact, regardless of the amount of missing data, this method always displays the highest normalised Robinson-Foulds score (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). When comparing this the probability of overlap in this method between the other onces, it is counter intuitive to note that the Bayesian consensus tree shows the highest dissimilarity with the Bayesian posterior trees from which the consensus tree is build (Fig.~\ref{Fig_Results-paircomp_within}-A). Note that it is also highly dissimilar to the Bootstraps. However that can be linked to the fact that both Bootstraps and Bayesian posterior trees are measured as 125*1000 random pairwise comparisons instead of just 125 tree comparisons for the Bayesian consensus trees. The fact that the 1000 trees to be compared for each of the 125 pairwise comparisons can add noise to the results for these two methods due to the random factor in the pairwise comparisons (see \hyperref[tree_comparisons]{Methods - Tree comparisons}).
